Deciding Bitvector Termination with Program Synthesis

### 2014-07-04 papers/termination/paper.tex
formalisation stuff

synthesizing environment abstractions: underapp for termination and underapp for non-termination.

\cite{} addresses the problem of automatically underapproximating weakest preconditions.

@@

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:ct}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

\noindent {\bf Conditional termination.}
If $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}.
The information we require of the environment is a predicate $I$ which
overapproximates the set of states that $L$ may begin executing in.
That is to say, for each state $x$ that is reachable on entry to $L$,
we have $I(x)$.

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T)$ terminates when started in any state satisfying $I(x)$ iff
 Defintion~\ref{def:ct} is satisfiable.
\end{theorem}

@@

\noindent {\bf Nested loops.}
If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
$L'$ in the sense that it satisfies the following formula:
\begin{align*}
 \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
\end{align*}
Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
the inner loop ($L'$)'s transition relation.

@@

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.  This figure contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

>>>>

@@

\begin{definition}[Conditional Termination Formula {\bf [CT]}]
\label{def:ct}
 \begin{align*}
  \exists R, W . \forall x, x' . & I(x) \wedge G(x) \rightarrow W(x) ~ \wedge \\
                                 & G(x) \wedge W(x) \wedge T_o(x, x') \rightarrow W(x') \wedge R(x) > 0 
  \wedge R(x) > R(x')
 \end{align*}
\end{definition}

@@

\noindent {\bf Conditional termination.}
Given a loop $L=(G,T_o)$, where $T_o$ is an over-approximation of its transition relation, 
if $L$'s termination depends on the state it begins
executing in, we say that $L$ is \emph{conditionally terminating}.
The information we require of the environment is a predicate $I$ which
overapproximates the set of states that $L$ may begin executing in.
That is to say, for each state $x$ that is reachable on entry to $L$,
we have $I(x)$.

@@

\begin{theorem}
\label{thm:ct}
 The loop $L(G, T_o)$ terminates when started in any state satisfying $I(x)$ iff
 Defintion~\ref{def:ct} is satisfiable.
\end{theorem}

@@

\noindent {\bf Environment abstractions.}
For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} defined in Definition\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
$I$ over-approximates the set of states that $L$ may begin executing in.
The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
is so that we can use the same formulae regardless of the environment the loop is executing in, or 
the complexity of its body, i.e. if it contains nested loops.

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any environment and in the presence of nested loops is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations.

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}...

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.

%synthesizing environment abstractions: underapp for termination and underapp for non-termination.
% \cite{} addresses the problem of automatically underapproximating weakest preconditions.

%\noindent {\bf Nested loops.}
If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
$L'$ in the sense that it satisfies the following formula:
\begin{align*}
 \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
\end{align*}
Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
the inner loop ($L'$)'s transition relation.

@@

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.

Figure~\ref{fig:environment-model} contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}
